Prediction rule ensembles in R

### 2016-08-30 pkg/glmertree/upre/paper/prediction_rule_ensembles.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@953 255f4249-2263-4417-b1ea-8498779e27a7

%% for \usepackage{Sweave}
\SweaveOpts{echo=FALSE, results=hide}
\setkeys{Gin}{width=0.8\textwidth}

<<preliminaries>>=
library("partykit")
@

@@

\begin{multline*}
	\\
	r_1(\mathbf{x}) = I (x_4 \leq 82)\\
	r_2(\mathbf{x}) = I (x_4 \leq 82 \cdot x_3 \in \{a,c\})\\
	r_3(\mathbf{x}) = I (x_4 \leq 82 \cdot x_3 \in \{b,d,e\})\\
	r_4(\mathbf{x}) = I (x_4 \leq 82 \cdot x_3 \in \{b,d,e\} \cdot x_4 \leq 77)\\
	r_5(\mathbf{x}) = I (x_4 \leq 82 \cdot x_3 \in \{b,d,e\} \cdot x_4 > 77)\\
	r_6(\mathbf{x}) = I (x_4 > 82)\\
	r_7(\mathbf{x}) = I (x_4 > 82 \cdot x_5 \leq \textrm{seldom})\\
	r_8(\mathbf{x}) = I (x_4 > 82 \cdot x_5 > \textrm{seldom})\\
\end{multline*}

@@

\begin{figure}[t!]
\setkeys{Gin}{width=\textwidth}
<<example-tree, fig=TRUE, width= 5, height= 5>>=
ex_tree <- party(
# node 0:
partynode(0L,
split = partysplit(2L, breaks = 82),
kids = list(
# node 1:
partynode(1L,
split = partysplit(1L, breaks = 2),
kids = list(
# node 2:
partynode(2L, info = ""), 
# node 3:
partynode(3L, 
split = partysplit(2L, breaks = 77),
kids = list(
# node 4:
partynode(4L, info = ""),
# node 5:
partynode(5L, info = ""))))),
# node 6:
partynode(6L,
split = partysplit(3L, breaks = 1),
kids = list(
# node 7:
partynode(7L, info = " "),
# node 8:
partynode(8L, info = " "))))), 
# dataset:
data.frame(x3 = factor(character(0), levels = c("a", "c", "b", "d", "e"), ordered = FALSE), 
x4 = numeric(0), 
x5 = factor(character(0), levels = c("seldom", "somtimes", "regularly"), ordered = TRUE))
)
plot(ex_tree, inner_panel = node_inner(ex_tree, id = FALSE), terminal_panel = node_terminal(ex_tree, id = FALSE, fill = "white"))
@
\label{fig:regression_tree}
\caption{Example conditional inference tree. Note that a conditional inference tree would normally depict the distribution of the outcome variable $y$ in each of the terminal nodes. For generating the initial ensemble of prediction rules, only the tree structure, not the distribution of the outcome variable is used, which is therefore omitted here.}
\end{figure}

@@

@@

Where $1 \dots j \dots n$ is an index for the input variable and $1 \dots k \dots K$ is an index for the rule in the initial ensemble. The coefficient values are estimated by means of penalized regression. When the final model is selected by lasso regression \cite{Tibs96}, the estimates satisfy the following minimization criterion:

@@

The first term is the prediction risk over the $N$ training observations, and the second term is a penalty for non-zero coefficients of rules and linear terms. The influence of the penalty term is determined by the value of $\lambda \geq 0$. The value of $\lambda$ is taken to be that which minimizes the prediction risk on a separate validation dataset, or by multi-fold cross validation on the training data. The penalty term in Equation \ref{eq:lasso_regression} forces many coefficients to be set to zero, providing a final ensemble consisting of a small set of rules and/or linear terms. Alternatively, ridge or elastic net penalties \cite{ZouyHast05} may be applied in Equation \ref{eq:lasso_regression}, instead of the lasso penalty, but this will generally yield less sparse models.

The penalty term in \ref{eq:lasso_regression} is sensitive to linear transformations of the input variables: generally, terms with smaller variance are more heavily penalized. Users can give a linear terms the same weight in the penalty term as a typical rule, by selecting its normalized values to be used in minimizing Equation \ref{eq:lasso_regression}:

\begin{equation} 
	l_j(x_j) \leftarrow 0.4*l_j(x_j) / \mathit{std}(l_j(x_j))
\end{equation}

Alternatively, all rules and linear functions may be standardized prior to estimating Equation \ref{eq:lasso_regression}, giving each rule and linear term equal a-priori influence. However, rules with support close to zero or one are defined by a correspondingly small number of training observations, and reducing penalization of these terms may be unwanted.

@@

@@

\begin{equation}
	I_j = |\hat{b}_j| \cdot \mathit{std}(l_j(x_j)) % equation 29 in F\&P
	\label{eq:global_linear_imp}
\end{equation}

@@

where $|S|$ is the cardinality (number of training set observations) in subregion $S$. Similarly, the local importance of a linear term over a selected subregion can be calculated by replacing the global standard deviation $\mathit{std}(l_j(x_j))$ in Equation \ref{eq:global_linear_imp} by its standard deviation in the selected subregion:

\begin{equation}
	\mathit{std}(l_j(x_j), S) = \sqrt{ \frac{1}{|S| - 1} \sum_{x_i \in S}{(l_j(x_ij) - \bar{l}_j(x_j, S))^2} }
	\label{eq:local_sd}
\end{equation}

@@

where $\bar{l}_j$ is the mean of $l_j(x_j)$ over the training data. The quantities in Equations \ref{eq:global_rule_imp}, \ref{eq:global_linear_imp}, \ref{eq:local_rule_imp} and \ref{eq:local_linear_imp} represent the absolute change in the (average) prediction $F(\mathbf{x})$ when the corresponding rule or linear term is dropped from the predictive model (Equation \ref{eq:predictive_model}), when the intercept would be adjusted accordingly (i.e., $\hat{a}_0 \leftarrow \hat{a}_0 + \hat{a}_k s_k$ for rules and $\hat{a}_0 \leftarrow \hat{a}_0 + \hat{b}_j \bar{l}_j$ for linear terms). The local importance of input variable $x_j$ at each point $\mathbf{x}$ is given by

@@

The influence of individual or pairs of predictor variables on the predictions of the final rule ensemble can be also assessed visually by plotting the partial dependence of $F(\mathbf{x})$ on individual or pairs of input variables. The partial dependence of $F(\mathbf{x})$ on a subset of input variable $s \subset \{1, ..., n\}$ is the expected value of $F(\mathbf{x})$ over the marginal joint distribution of variables $\mathbf{x}_{\backslash s}$ not in $\mathbf{x}_s$. It is defined as

\begin{equation}
	F_s(\mathbf{x}_s) = E_{\mathbf{x}_{\backslash s}} [F(\mathbf{x}_s, \mathbf{x}_{\backslash s})] % equation 39 in F/&P
\end{equation}

and can be estimated from the data by

\begin{equation}
	\hat{F}_s(\mathbf{x}_s) = \frac{1}{N} \sum^N_{i=1}{F(\mathbf{x}_s, \mathbf{x}_{i \backslash s})} % equation 40 in F/&P
\end{equation}

where $\{ \mathbf{x}_{i \backslash s} \}^N_1$ are the observations in the training dataset. The subset of input variables $s$ must consist of single or a pair of input variables to allow for plotting the partial dependence of $F(\mathbf{x})$ on $\mathbf{x}_s$.

@@

Prediction rules are well suited to capturing interaction effects of input variables. However, the presence of prediction rules involving multiple variables in the final ensemble are a necessary but not sufficient condition for the presence of interaction effects. Different rules involving the same sets of variables may eliminate the interaction effect of these variable in the final predictive model $F(\mathbf{x})$. Therefore, the presence of interaction effects should be assessed by analyzing the full predictive model $F(\mathbf{x})$, for example by means of partial dependence plots and computation of interaction statistics, which will be discusses in this section.

If an input variable $x_j$ does not interact with any of the other input variables, then the effect of $x_j$ and $x_{\backslash j}$ on $F(\mathbf{x})$ are additive, and therefore can be expressed as

\begin{equation}
	F(\mathbf{x}) = F_j(x_j) + F(\mathbf{x}_{\backslash j}) % equation 42
\end{equation}

where $F_j(x_j)$ is the partial dependence of $F(\mathbf{x})$ on $x_j$ and $F(\mathbf{x}_{\backslash j})$ is the partial dependence of $F(\mathbf{x})$ on $\mathbf{x}_{\backslash j}$. A statistics for testing whether $x_j$ interacts with any other input variable is given by

\begin{equation}
	H^2_j = \frac{\sum^N_{i=1}{ [F(\mathbf{x}_i) - \hat{F}_j(x_{ij}) - \hat{F}_{\backslash j}(x_{i \backslash j})]^2 }}{{\sum^N_{i=1}{ F^2(\mathbf{x}_i)}
	\label{eq:interaction_statistic}
\end{equation}

$H^2_j$ will differ from zero to the extent that $x_j$ is involved in interactions with other input variables. To assess whether $H^2_j$ differs significantly from zero, a null distribution for the statistic has to be derived. \cite{FrieyPope08} suggest the use of a variant of the parametric bootstrap \cite{EfroyTibs94} to derive a null distribution of statistic $ H_j^2$. The statistic $H^2_j$ is repeatedly computed on a series of artificial datasets from which interactions involving $x_j$ are absent. The distribution of test statistics thus obtained can be used as a reference distribution, to assess significance of the value of  $H^2_j$ based on the training data.

%To disencourage spurious interactions entering the model, F\&P (section 8.2) use a strategy to put an incentive on variables entering a tree, that do not appear earlier in the tree. I currently do not know of a way to do that for ctrees.

@@

- ctree instead of CART trees
- Rulefit uses an approach that can be best described as a combination of bagging and boosting: it takes subsamples\footnote{With subsampling, observations are drawn without replacement from the full dataset, so strictly speaking, Rulefit does not use a bagging approach, as this would involve bootstrap samples of the dataset.} of size $\eta$ of the training data, and grows trees in a sequential fashion with learning rate $\nu = 0, \dots, 1$.
- Rulefit allows for inducing trees in a manner that can be best described as a hybrid of bagging \cite{Brei96bagging} and boosting \cite{Frie01}. In some situations, bagged ensembles of trees may be outperformed by random forests in terms of predictive accuracy \cite{Diet00bbrf}. Also, Node Harvest derives prediction rules from the nodes of random forest trees. Therefore, upre allows for inducing trees in a bagging, boosting or random forest type manner, or a hybrid of the three.    
- Rulefit also provides forward stagewise regression
- Current implementation of rulefit does not provide user-friendly interface for rules inspection, does not provide value of lambda or intercept, and does not allow user to select a different values for lambda (e.g., lamba min or lambda 1se).
- In the pairplot function, rulefit does not seem to cross all possible values of x1 and x2, it may just take all observed combinations. Then, correlated x1 and x2 may result in a plot that looks like an interaction, but it is not. Check if my plotting function gives a different results, although it may be slower (but then, it is immediately clear why))

@@

Users are adviced to not use the lambda path determined by \verb|glmnet|, include a way to determine own path?

Future implementations may support multivariate outcomes. ctree() allows for multivariate responses. If glmnet also allows for that, multivariate rule ensembles can be created.

\subsection{Extensions to random-effect models}

Schelldorder, Meier \& Buhlmann (2014) propose a two-step approach for high-dimensional GLMMs: In the first step, variable screening is performed, selecting a set of candidate active variables. In the second step, refitting by maximum likelihood (ML) estimation is performed to get accurate parameter estimates.
Groll \& Tutz (2010) propose a method that works by combining gradient ascent optimization with the Fisher scoring algorithm and is based on the approach of Goeman (2010).

>>>>

%% for \usepackage{Sweave}

@@

\begin{multline*}
	r_1(\mathbf{x}) = I (x_4 \leq 82)\\
	r_2(\mathbf{x}) = I (x_4 \leq 82 \cdot x_3 \in \{a,c\})\\
	r_3(\mathbf{x}) = I (x_4 \leq 82 \cdot x_3 \in \{b,d,e\})\\
	r_4(\mathbf{x}) = I (x_4 \leq 82 \cdot x_3 \in \{b,d,e\} \cdot x_4 \leq 77)\\
	r_5(\mathbf{x}) = I (x_4 \leq 82 \cdot x_3 \in \{b,d,e\} \cdot x_4 > 77)\\
	r_6(\mathbf{x}) = I (x_4 > 82)\\
	r_7(\mathbf{x}) = I (x_4 > 82 \cdot x_5 \leq \textrm{seldom})\\
	r_8(\mathbf{x}) = I (x_4 > 82 \cdot x_5 > \textrm{seldom})\\
\end{multline*}

@@

\begin{figure}[t!]
\setkeys{Gin}{width=\textwidth}
\includegraphics{prediction_rule_ensembles-example-tree}
\label{fig:regression_tree}
\caption{Example conditional inference tree. Note that a conditional inference tree would normally depict the distribution of the outcome variable $y$ in each of the terminal nodes. For generating the initial ensemble of prediction rules, only the tree structure, not the distribution of the outcome variable is used, which is therefore omitted here.}
\end{figure}

@@

% !!! rules that are each others negative should also be removed!!!!

@@

Where $1 \dots j \dots n$ is an index for the input variable and $1 \dots k \dots K$ is an index for the rule in the initial ensemble. The coefficient values are estimated by means of penalized regression. When the final model is selected by lasso regression \cite{Tibs96}, which is the default in the upre algorithm, the estimates satisfy the following minimization criterion:

@@

The first term is the prediction risk over the $N$ training observations, and the second term is a penalty for non-zero coefficients of rules and linear terms. The influence of the penalty term is determined by the value of $\lambda \geq 0$. The value of $\lambda$ is taken to be that which minimizes the prediction risk on a separate validation dataset, or by multi-fold cross validation on the training data. The penalty term in \ref{eq:lasso_regression} forces many coefficients to be set to zero, providing a final ensemble consisting of a small set of rules and/or linear terms. Users can also select ridge or elastic net regression \cite{ZouyHast05} to derive the final ensemble. The penalty term in Equation \ref{eq:lasso_regression} would then replaced by a ridge or elastic net counterpart, respectively penalizing squared values of $\hat{a}_k$ and $\hat{b}_j$ instead, or as well as, absolute values.

The penalty term in \ref{eq:lasso_regression} is affected by linear transformations of the input variables: generally, terms with smaller variance are more heavily penalized. Users can give linear terms the same influence on the penalty term as a typical rule, by selecting their normalized versions to be used in the estimation (Equation \ref{eq:lasso_regression}):

\begin{equation} 
	l_j(x_j) <- 0.4*l_j(x_j) / \textit{std}(l_j(x_j))
\end{equation}

% Rules are not normalized. Rules with support close to zero or one are defined by a correspondingly small number of training observations, and therefore it is OK that lasso regression puts increased penalty on those.

@@

% if linear terms are normalized, the intercept should be corrected in the coef and/or predict function, as well.

@@

\begin{equation}
	I_j = |\hat{b}_j| \cdot \textit{std}(l_j(x_j)) % equation 29 in F\&P
	\label{eq:global_linear_imp}
\end{equation}

@@

where $|S|$ is the cardinality (number of training set observations) in subregion $S$. Similarly, the local importance of a linear term over a selected subregion can be calculated by replacing the global standard deviation $\textit{std}(l_j(x_j))$ in Equation \ref{eq:global_linear_imp} by its standard deviation in the selected subregion:

\begin{equation}
	\textit{std}(l_j(x_j), S) = \sqrt{ \frac{1}{|S| - 1} \sum_{x_i \in S}{(l_j(x_ij) - \bar{l}_j(x_j, S))^2} }
	\label{eq:local_sd}
\end{equation}

@@

where $\bar{l}_j$ is the mean of $l_j(x_j)$ over the training data. The quantities in Equations \ref{eq:global_rule_imp}, \ref{eq:global_linear_imp}, \ref{eq:local_rule_imp} and \ref{eq:local_linear_imp} represent the absolute change in the (average) prediction $F(\mathbf{x})$ when the corresponding rule or linear term is dropped from the predictive model (Equation \ref{eq:predictive_model}), when the intercept would be adjusted accordingly (i.e., $\hat{a}_0 <- \hat{a}_0 + \hat{a}_k s_k$ for rules and $\hat{a}_0 <- \hat{a}_0 + \hat{b}_j \bar{l}_j$ for linear terms). The local importance of input variable $x_j$ at each point $\mathbf{x}$ is given by

@@

The influence of individual predictor variables on the predictions of the final rule ensemble can also be represented visually in partial dependence plots

@@

One of the main advantages of tree-based methods is the automatic detection of non-linear and/or interaction effects. However, the presence of interactions or non-linear effects of single variables cannot simply be inferred from inspection of the rules in the final ensemble. On the one hand, variables involved in interaction effects do not necessarily appear in the same rules, and on the other hand, variables appearing in the same rules do not necessarily produce an interaction effect on the outcome variable. To assess non-linear effects and multi-way interactions of predictor variables, partial dependence plots and statistical tests are provided.

In addition, the presence of interactions of (user-specified subsets of) predictor variables can be statistically tested. For these tests, a null model involving main effects of predictor variables only, is compared with a model involving interactions.

To disencourage spurious interactions entering the model, F\&P (section 8.2) use a strategy to put an incentive on variables entering a tree, that do not appear earlier in the tree. Meinshausen uses a random forest methodology for deriving the trees for the initial ensemble, which may have the same results: by selecting only a subset of the predictor variables for growing each tree, interaction effects will be disencouraged.

@@

- ctree instead of CART trees
- Rulefit uses an approach that can be best described as a combination of bagging and boosting: it takes subsamples\footnote{With subsampling, observations are drawn without replacement from the full dataset, so strictly speaking, Rulefit does not use a bagging approach, as this would involve bootstrap samples of the dataset.} of size $\eta$ of the training data, and grows trees in a sequential fashion with learning rate $\nu = 0, \dots, 1$.
- Rulefit allows for inducing trees in a manner that can be best described as a hybrid of bagging \cite{Brei96bagging} and boosting \cite{Frie01}. In some situations, bagged ensembles of trees may be outperformed by random forests in terms of predictive accuracy \cite{Diet00bbrf}. Also, Node Harvest derives prediction rules from the nodes of random forest trees. Therefore, upre allows for inducing trees in a bagging, boosting or random forest type manner, or a hybrid of the three.    
- Rulefit also provides forward stagewise regression
- Current implementation of rulefit does not provide user-friendly interface for rules inspection, does not provide value of lambda or intercept, and does not allow user to select a different values for lambda (e.g., lamba min or lambda 1se).

@@

Users are advices to not use the lambda path determined by \verb|glmnet|, include a way to determine own path?

\subsection{Extensions to random-effect models}
Schelldorder, Meier \& Buhlmann (2014) propose a two-step approach for high-dimensional GLMMs: In the first step, variable screening is performed, selecting a set of candidate active variables. In the second step, refitting by maximum likelihood (ML) estimation is performed to get accurate parameter estimates.
Groll \& Tutz (2010) propose a method that works by combining gradient ascent optimization with the Fisher scoring algorithm and is based on the approach of Goeman (2010).
